Association between trauma severity models and opportunities for improvement:

### 2022-09-19 studyprogressreport.Rmd

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

## Trauma severity scores
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity scores (TSS). TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).

@@

@@

<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? -->

@@

Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->

@@

<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. Do I need to have the formulas? -->

@@

>>>>

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic<!-- I typically define trauma as the clinical entity composed of physical injury and the body's associated response. I would suggest that you stick to trauma and don't use physical injury and trauma interchangeably -->. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

## Trauma severity scores
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity scores (TSS)<!-- Avoid non-standard abbreviations, as it makes your text harder to read. For example, I don't think spelling out trauma severity scores is such a big deal, especially considering that you'll be dealing with so many score names that are abbreviations themselves -->. TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).

@@

<!-- Note that you mix trauma severity scores, trauma severity models, prediction models, etc. Stick to one of these concepts -->

@@

<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? --> <!-- That's fine -->

@@

Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> <!-- Don't include numbers here, those belongs in Reults -->

@@

<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. Do I need to have the formulas? --> <!-- Yes, you need the formulas, but not for this report -->

@@

<!-- The aim includes both to determine association between the severity estimates and OFI as well as how well they predict OFI. These two components require different methods. To determine the strength of the association you could use logistic regression, then the resulting OR would be your effect size. Because all these models returns a probability you don't need to standardise the scores to be able to compare the ORs across models to determine which score that is most strongly associated with OFI. To assess how well they predict OFI you need to calculate the discrimination (often expressed as the area under the receiver operating characteristics curve, or just AUC) and calibration (using for example the integrated calibration index. Note that this is slightly contrived, as that would equate the probability of death with the probability of OFI-->
